---
title: "project"
output: html_document
date: "2024-03-29"
---

```{r setup, include=FALSE}
library(ggplot2)
```

```{r}
datapr<-read.csv("Wholesale.csv", header = TRUE, sep = ",")#import data
datapca <- datapr
datapca$Channel <- NULL  # Remove the 'Channel' column/ 
datapca$Region <- NULL 
names(datapca)

```

```{r}
ts.plot(data,gpars= list(col=rainbow(20)))#plot series
legend("topleft", legend = colnames(datapca), col=rainbow(20), lty = 1,cex = 0.5 )

data_s<-datapca
data_s<- scale(datapca)#standardize data
ts.plot(data_s,gpars= list(col=rainbow(20)))
legend("topleft", legend = colnames(data_s), col=rainbow(20), lty = 1,cex = 0.5)

```

```{r}

cov.m<-cov(datapca)# compute coovariance matrix
cov.m
round(cov.m,digit=2) #see only two digits after the comma 
max(diag(cov.m))
min(diag(cov.m))
cor.m<-cor(datapca)#compute the correlation matrix
round(cor.m,digit=2)
?princomp
pca<-princomp(datapca,cor=TRUE) #principal components analysis on the correlation matrix
pca

allpca <-princomp(datapca,cor=FALSE) 
allpca
```
# Table in Latex (correlation matrix)
```{r}
library(xtable)

# Your correlation matrix 'cor_matrix'
cor_matrix <- matrix(c(1, 0.1, -0.01, 0.35, -0.1, 0.24,
                       0.1, 1, 0.73, 0.12, 0.66, 0.41,
                       -0.01, 0.73, 1, -0.04, 0.92, 0.21,
                       0.35, 0.12, -0.04, 1, -0.13, 0.39,
                       -0.1, 0.66, 0.92, -0.13, 1, 0.07,
                       0.24, 0.41, 0.21, 0.39, 0.07, 1), byrow = TRUE, ncol = 6)

rownames(cor_matrix) <- colnames(cor_matrix) <- c("Fresh", "Milk", "Grocery", "Frozen", "Detergents_Paper", "Delicassen")

cor_matrix[upper.tri(cor_matrix)] <- ""
cor_df <- as.data.frame(cor_matrix)
latex_table <- print(xtable(cor_df), include.rownames = TRUE, include.colnames = TRUE, comment = FALSE)

```
# We calculate eigen value and the proportion of variance (based on correlation matrix)

```{r}
screeplot(pca, npcs =17,type = c("lines"),ylim=c(0,5))
summary(pca)
pr.var <- pca$sdev^2#eigen values
pr.variance <- pca$sdev^2

pve <- pr.var / sum(pr.var)
pr.var
cumsum = cumsum (pve)* 100
frame  = data.frame(pr.var, pve, cumsum)
frame
plot(cumsum(pve[1:17]), xlab = "Principal Component",ylab = "Proportion of Variance Explained",xlim=c(0,6),ylim = c(0, 1), type = "b")
```

# We calculate eigen value and the proportion of variance (based on covariance matrix)
```{r}
screeplot(allpca, npcs =17,type = c("lines"),ylim=c(0,5))
summary(allpca)
pr.var <- allpca$sdev^2#eigen values
pr.variance <- pca$sdev^2

pve <- pr.var / sum(pr.var)
pr.var
cumsum = cumsum (pve)* 100
frame  = data.frame(pr.var, pve, cumsum)
frame
plot(cumsum(pve[1:17]), xlab = "Principal Component",ylab = "Proportion of Variance Explained",xlim=c(0,6),ylim = c(0, 1), type = "b")
```
# Loadings

```{r}
pca2<-prcomp(datapca)
pca2
load <- data.frame(Category = rownames(pca2$rotation), pca2$rotation[, 1:3])
load
colnames(load) = c("Category", "PC1","PC2", "PC3")
load

```
PC1 Interpretation 

The 'Fresh' category has a large negative loading on PC1 (-0.9765), which means that PC1 largely represents the spending on fresh products. Clients further to the left on the PC1 axis on your scatter plot tend to spend more on fresh products, while those further to the right tend to spend less.
Other categories like 'Milk', 'Grocery', and 'Detergents_Paper' have smaller negative or positive loadings on PC1, indicating that they have a less significant but still meaningful contribution to the variation captured by PC1.

PC2 Interpretation

The 'Grocery' category has a high positive loading on PC2 (0.7646), suggesting that PC2 captures variance associated with grocery spending. Clients higher on the PC2 axis spend more on grocery products.
'Milk' and 'Detergents_Paper' also contribute positively to PC2, albeit to a lesser extent. This suggests that clients with high values on PC2 also tend to spend more on milk and detergents/paper products.
The 'Fresh' category has a small negative loading on PC2, indicating that it does not influence PC2 as much as PC1.

# Plot to interpret components PC1 and PC2
```{r}
getwd()

ggplot(data = load, aes(x = PC1, y = PC2, label = Category)) +
  geom_point() +  # This adds the points to the plot
  geom_text(aes(label=Category), hjust=0.5, vjust=-0.5, color = "blue") +  # This adds the text labels in blue
  scale_x_continuous(limits = c(-1, 1), breaks = seq(-1, 1, 0.5)) +  # Set x-axis limits and breaks
  scale_y_continuous(limits = c(-1, 1), breaks = seq(-1, 1, 0.5)) +  # Set y-axis limits and breaks
  geom_vline(xintercept = 0, linetype="dashed", color = "black") +  # Add a vertical line at x=0
  geom_hline(yintercept = 0, linetype="dashed", color = "black") +  # Add a horizontal line at y=0
  labs(title = "PCA Loadings Plot", x = "Principal Component 1 (PC1)", y = "Principal Component 2 (PC2)") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1),  # Rotate x-axis text if needed
        plot.title = element_text(hjust = 0.5))  # Center the title
# Save the plot
ggsave("pca_loadings_plot.pdf", width = 7, height = 5, device = "pdf")
```

```{r}
datatr=t(datapca)
pca1<-prcomp(datatr) #pca analysis
pca1$rotation[,1:2]#loadings of the first two components
pc = data.frame(pca1$rotation[,1:3])
pc$Region = datapr$Region
pc$Channel = datapr$Channel
```
# Plot differences beetwenn regions 

Here are no visible clusters, probably will not use them.
```{r}
ggplot(pc, aes(x = PC1, y = PC2, color = as.factor(Region))) +
  geom_point() +
  theme_minimal() +
  labs(color = "Region")

ggplot(pc, aes(x = PC1, y = PC3, color = as.factor(Region))) +
  geom_point() +
  theme_minimal() +
  labs(color = "Region")
ggplot(pc, aes(x = PC2, y = PC3, color = as.factor(Region))) +
  geom_point() +
  theme_minimal() +
  labs(color = "Region")
```

# Plot differences between channels

```{r}
pc$Channel <- factor(pc$Channel, levels = c(1, 2), labels = c("Horeca", "Retail"))


p1= ggplot(pc, aes(x = PC1, y = PC2, color = as.factor(Channel))) +
  geom_point() +
  scale_color_manual(values = c("Horeca" ="#E41A1C", "Retail" ="#377EB8")) + # Change point colors
  theme_minimal() +
  labs(color = "Channel")
ggsave("PC1_vs_PC2.pdf", p1, width = 7, height = 7, device = "pdf")


p2=ggplot(pc, aes(x = PC1, y = PC3, color = as.factor(Channel))) +
  geom_point() +
   scale_color_manual(values = c("Horeca"="#4DAF4A",  "Retail" ="#984EA3")) + # Change point colors
  theme_minimal() +
  labs(color = "Channel")
ggsave("PC1_vs_PC3.pdf", p2, width = 7, height = 7, device = "pdf")


p3=ggplot(pc, aes(x = PC2, y = PC3, color = as.factor(Channel))) +
  geom_point() +
   scale_color_manual(values = c("Horeca"="#FF7F00","Retail"= "#FFFF33")) + # Change point colors
  theme_minimal() +
  labs(color = "Channel")


ggsave("PC2_vs_PC3.pdf", p3, width = 7, height = 7, device = "pdf")

p1
p2
p3
```

Realtionship between PC1 and PC2 (graph1)

Horeca clients (Channel 1) are clustered to the left, they might be characterized by higher spending on fresh products, while Retail clients (Channel 2) might be represented by higher spending on groceries and other non-fresh products.


# Plot from lab
```{r}
plot(pca1$rotation[,1],xlab="Costumers")
abline(h=0)
plot(pca1$rotation[,2],xlab="Costumers")
abline(h=0)
```


```{r}
datapr$Region=as.factor(datapr$Region)

#churn rate in the dataset
prop.table((table(datapr$Region)))

```

```{r}
#boxplot
?boxplot
boxplot(datapr$Fresh~datapr$Region)
boxplot(datapr$Milk~datapr$Region)
boxplot(datapr$Grocery~datapr$Region)
boxplot(datapr$Frozen~datapr$Region)
boxplot(datapr$Detergents_Paper~datapr$Region)
boxplot(datapr$Delicassen~datapr$Region)

```
```{r}
data_s<-datapr
#standardize data
data_s[,1:(ncol(data_s)-1)]<-scale(datapr[,1:(ncol(datapr)-1)])
library(caTools)
set.seed(123)
split = sample.split(data_s$Region, SplitRatio = 0.7)
train= subset(data_s, split == TRUE)
test= subset(data_s, split == FALSE)
table(train$Region)
table(test$Region)
```
```{r}
library(MASS)
model_LDA<-lda(Region~.,data=train)

```

```{r}
library(biotools)
pred_train<-predict(model_LDA,data=train)$class
ct_train<-confusionmatrix(train$Region, pred_train)
sum(diag(prop.table(ct_train)))
#classification accuracy 86%
```

```{r}
6.
pred_test<-predict(model_LDA, test)$class
ct_test<-confusionmatrix(test$Region, pred_test)
sum(diag(prop.table(ct_test)))
#classification accuracy 86%
```

```{r}
model_QDA<-qda(Region~.,data=train)
pred_train_QDA<-predict(model_QDA,data=train)$class
ct_train_QDA<-confusionmatrix(train$region, pred_train_QDA)
sum(diag(prop.table(ct_train_QDA)))
```



```{r}

pred_test_QDA<-predict(model_QDA, test)$class
ct_test_QDA<-confusionmatrix(test$Region, pred_test_QDA)
sum(diag(prop.table(ct_test_QDA)))
#classification accuracy 89%
```


7.





